A Linear Solution of the Linearity Challenge

### 2025-12-12 paper/Congruence.lagda.tex
revised congruence

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)

@@

\begin{AgdaAlign}
\begin{code}
data _⊒_ {Γ} : Process Γ → Process Γ → Set where
\end{code}

@@

\begin{code}
  s-comm :
    ∀{A Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
    cut {A} p P Q ⊒ cut (+-comm p) Q P
\end{code}

The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
property of parallel composition. We use \AgdaFunction{+-comm} to compute the
proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$
respectively.
%
Notice that this constructor also makes key use of the implicit rewriting rule
\AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
have type $\AgdaDatatype{Process}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$
and
$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
respectively, but the \AgdaInductiveConstructor{cut} on the r.h.s. of
\AgdaInductiveConstructor{⊒} expects $P$ to have type
$\AgdaFunction{dual}~(\AgdaFunction{dual}~A)$. Thanks to
\AgdaFunction{dual-inv}, Agda considers these types equivalent without requiring
intricate substitutions in the index of a data type.

\begin{code}[hide]
  s-link :
    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) → link p ⊒ link (+-comm p)
  s-fail :
    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊤ ⊳ Δ) →
    let _ , _ , q′ = +-assoc-l p q in
    cut {A} p (fail (> q)) P ⊒ fail q′
\end{code}

@@

\begin{code}
  s-wait :
    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊥ ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (wait (> q) P) Q ⊒ wait q′ (cut p′ P Q)
\end{code}

There are two non-trivial aspects worth commenting. The first one concerns the
proof $(\AgdaInductiveConstructor{>}~q)$ used in the
\AgdaInductiveConstructor{wait} process before structural precongruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{>}~q)~P)$ is a
    direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore it
    is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
    %
    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
    from the right component of the splitting $[\Bot] +
    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
    \AgdaInductiveConstructor{>}.
\end{enumerate}

The other aspect that is worth commenting concerns the rearrangement of the
splittings in the process after the application of structural precongruence.
Overall, $p$ and $q$ prove the splittings $([\Bot] + \ContextD) + \ContextC_2$,
but the precongruence rule requires this splitting to be rearranged as $[\Bot] +
(\ContextD + \ContextC_2)$. That is, we need to apply the left-to-right
associativity property of context splitting which we called
\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}. The nested
\AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern match the result of the
application $\AgdaFunction{+-assoc-l}~p~q$ and to extract the new proofs $p'$
and $q'$ for the rearranged splittings.

The constructors \AgdaInductiveConstructor{s-left} and
\AgdaInductiveConstructor{s-right} model \SSelect when the selected tag is
respectively $\Tag_1$ and $\Tag_2$.

\begin{code}[hide]
  s-case :
    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B & C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (case (> q) P Q) R ⊒ case q′ (cut (< p′) (↭process swap P) R)
                                           (cut (< p′) (↭process swap Q) R)
\end{code}
\begin{code}
  s-left :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (left (> q) P) Q ⊒ left q′ (cut (< p′) (↭process swap P) Q)
\end{code}
\begin{code}[hide]
  s-right :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (right (> q) P) Q ⊒ right q′ (cut (< p′) (↭process swap P) Q)
\end{code}

Here the process
$(\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{>}~q)~P)$, which is
found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
tag $\Tag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ will be using a fresh continuation channel $z :
B$. Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type of the continuation sits on top of the type of the restricted channel.
%
After structural precongruence is applied, however, the type $B$ of the
continuation channel $z$ ends up underneath that of the restricted channel $x$,
because now the two channels are introduced in the opposite order. Therefore, we
are in a situation where we need to turn $P$ into a process that is well typed
in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. This
is achieved applying the function \AgdaFunction{↭process} to the
\AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
in a typing context, and finally to the process $P$.

We also show the modeling of the \SForkL rule, which is interesting because of
its complex side conditions:

\begin{code}[hide]
  s-join :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⅋ C ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (join (> q) P) Q ⊒
    join q′ (cut (< < p′) (↭process (↭shift {A} {C ∷ B ∷ []}) P) Q)
\end{code}
\begin{code}
  s-fork-l :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
    cut {A} p (fork (> q) (< r) P Q) R ⊒
    fork q′ r′′ (cut (< q′′) (↭process swap P) R) Q
\end{code}

@@

\begin{code}[hide]
  s-fork-r :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    cut {A} p (fork (> q) (> r) P Q) R ⊒
    fork q′ r′ P (cut (< p′′) (↭process swap Q) R)
  s-all :
    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∀ B ⊳ Δ) ->
    let _ , p' , q' = +-assoc-l p q in
    cut {A} p (all (> q) F) Q ⊒ all q' λ σ → cut (< p') (↭process swap (F σ)) Q
  s-ex :
    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∃ B ⊳ Δ) ->
    let _ , p' , q' = +-assoc-l p q in
    cut {A} p (ex {_} {C} (> q) P) Q ⊒ ex q' (cut (< p') (↭process swap P) Q)
  s-server :
    ∀{A B Γ₁ Γ₂ Δ₁ P Q}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `! B ⊳ Δ₁) (r : Γ₂ ≃ [] + Γ₂)
    (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {`? A} p (server (> q) (un-∷ un₁) P) (server (< r) un₂ Q) ⊒
    server q′ (+-un p′ un₁ un₂) (cut (< p′) (↭process swap P) (server (< r) un₂ Q))
  s-client :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (client (> q) P) Q ⊒ client q′ (cut (< p′) (↭process swap P) Q)
  s-weaken :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (weaken (> q) P) Q ⊒ weaken q′ (cut p′ P Q)
  s-contract :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} p (contract (> q) P) Q ⊒
    contract q′ (cut (< < p′) (↭process (↭shift {A} {`? B ∷ `? B ∷ []}) P) Q)
\end{code}

@@

\begin{code}
  s-refl  : ∀{P} → P ⊒ P
  s-tran  : ∀{P Q R} → P ⊒ Q → Q ⊒ R → P ⊒ R
  s-cong  : ∀{Γ₁ Γ₂ A P Q P′ Q′} (p : Γ ≃ Γ₁ + Γ₂) →
            P ⊒ Q → P′ ⊒ Q′ → cut {A} p P P′ ⊒ cut p Q Q′
\end{code}

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Unit using (tt)
open import Data.Sum using (inj₁; inj₂)
open import Data.Product using (_,_)
open import Data.List.Base using ([]; _∷_; [_])

@@

\begin{AgdaAlign}
\begin{code}
data _⊒_ {Γ} : Proc Γ → Proc Γ → Set where
\end{code}

@@

\begin{code}
  s-comm :
    ∀{A Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
    cut {A} (P ⟨ p ⟩ Q) ⊒ cut (Q ⟨ +-comm p ⟩ P)
\end{code}

The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
property of parallel composition. We use \AgdaFunction{+-comm} to compute the
proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$.
%
Notice that this constructor also makes key use of the implicit rewriting rule
\AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
have type $\AgdaDatatype{Process}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$
and
$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
respectively, but the \AgdaInductiveConstructor{cut} on the r.h.s. of
\AgdaInductiveConstructor{⊒} expects $P$ to have type
$\AgdaFunction{dual}~(\AgdaFunction{dual}~A)$. Thanks to
\AgdaFunction{dual-inv}, Agda considers these types equivalent without requiring
intricate substitutions in the index of a data type.

\begin{code}[hide]
  s-link :
    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) →
    link (ch ⟨ p ⟩ ch) ⊒ link (ch ⟨ +-comm p ⟩ ch)
  s-fail :
    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ ⊤ ] + Δ) →
    let _ , _ , q′ = +-assoc-l p q in
    cut {A} (fail (ch ⟨ > q ⟩ tt) ⟨ p ⟩ P) ⊒ fail (ch ⟨ q′ ⟩ tt)
\end{code}

@@

\begin{code}
  s-wait :
    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ ⊥ ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (wait (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
    wait (ch ⟨ q′ ⟩ cut (P ⟨ p′ ⟩ Q))
\end{code}

There are two non-trivial aspects worth commenting. The first one concerns the
proof $(\AgdaInductiveConstructor{>}~q)$ used in the
\AgdaInductiveConstructor{wait} process before structural precongruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
    %
    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
    from the right component of the splitting $[\Bot] +
    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
    \AgdaInductiveConstructor{>}.
\end{enumerate}

The other aspect that is worth commenting concerns the rearrangement of the
splittings in the process after the application of structural precongruence.
Overall, $p$ and $q$ prove the splittings $([\Bot] + \ContextD) + \ContextC_2$,
but the precongruence rule requires this splitting to be rearranged as $[\Bot] +
(\ContextD + \ContextC_2)$. That is, we need to apply the left-to-right
associativity property of context splitting which we called
\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}.\Luca{Controllare
left/right.} The nested \AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern
match the result of the application $\AgdaFunction{+-assoc-l}~p~q$ and to
extract the new proofs $p'$ and $q'$ for the rearranged splittings.

The constructors \AgdaInductiveConstructor{s-select-l} and
\AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
respectively $\InTag_1$ and $\InTag_2$.\Luca{Controllare nomi dei tag con
sintassi del calcolo. Chiamarli inj come in Agda così c'è consistenza?}
%
For example, for \AgdaInductiveConstructor{s-select-l} we have:

\begin{code}[hide]
  s-case :
    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B & C ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (case (ch ⟨ > q ⟩ (P , Q)) ⟨ p ⟩ R) ⊒
    case (ch ⟨ q′ ⟩ (cut (↭proc swap P ⟨ < p′ ⟩ R) ,
                     cut (↭proc swap Q ⟨ < p′ ⟩ R)))
\end{code}
\begin{code}
  s-select-l :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊕ C ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (select (ch ⟨ > q ⟩ inj₁ P) ⟨ p ⟩ Q) ⊒
    select (ch ⟨ q′ ⟩ inj₁ (cut (↭proc swap P ⟨ < p′ ⟩ Q)))
\end{code}
\begin{code}[hide]
  s-select-r :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊕ C ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (select (ch ⟨ > q ⟩ inj₂ P) ⟨ p ⟩ Q) ⊒
    select (ch ⟨ q′ ⟩ inj₂ (cut (↭proc swap P ⟨ < p′ ⟩ Q)))
\end{code}

Here the process
$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, which is
found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
tag $\InTag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ uses a fresh continuation channel $z : B$.
Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type of the continuation sits on top of the type of the restricted channel.
%
After structural precongruence is applied, however, the type $B$ of the
continuation channel $z$ ends up underneath that of the restricted channel $x$,
because now the two channels are introduced in the opposite order. Therefore, we
need to rename the channels in $P$ so that it is well typed in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
renaming is achieved applying the function \AgdaFunction{↭proc} to the
\AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
in a typing context, and finally to the process $P$.

\begin{code}[hide]
  s-join :
    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⅋ C ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (join (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
    join (ch ⟨ q′ ⟩ cut (↭proc (↭shift {A} {C ∷ B ∷ []}) P ⟨ < < p′ ⟩ Q))
\end{code}

We also discuss the modeling of the \SForkL rule, which is interesting because of
its complex side conditions:

\begin{code}
  s-fork-l :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊗ C ] + Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
    cut {A} (fork (ch ⟨ > q ⟩ (P ⟨ < r ⟩ Q)) ⟨ p ⟩ R) ⊒
    fork (ch ⟨ q′ ⟩ (cut (↭proc swap P ⟨ < q′′ ⟩ R) ⟨ r′′ ⟩ Q))
\end{code}

@@

\begin{code}[hide]
  s-fork-r :
    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊗ C ] + Δ) (r : Δ ≃ Δ₁ + Δ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    let _ , p′′ , r′ = +-assoc-l p′ r in
    cut {A} (fork (ch ⟨ > q ⟩ (P ⟨ > r ⟩ Q)) ⟨ p ⟩ R) ⊒
    fork (ch ⟨ q′ ⟩ (P ⟨ r′ ⟩ cut (↭proc swap Q ⟨ < p′′ ⟩ R)))
  s-all :
    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `∀ B ] + Δ) ->
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (all (ch ⟨ > q ⟩ F) ⟨ p ⟩ Q) ⊒
    all (ch ⟨ q′ ⟩ λ X → cut (↭proc swap (F X) ⟨ (< p′) ⟩ Q))
  s-ex :
    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `∃ B ] + Δ) ->
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (ex {_} {C} (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
    ex (ch ⟨ q′ ⟩ cut (↭proc swap P ⟨ < p′ ⟩ Q))
  s-server :
    ∀{A B Γ₁ Γ₂ Δ₁ P Q}
    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `! B ] + Δ₁) (r : Γ₂ ≃ [] + Γ₂)
    (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {`? A} (server (ch ⟨ > q ⟩ (un-∷ un₁ , P)) ⟨ p ⟩ server (ch ⟨ < r ⟩ (un₂ , Q))) ⊒
    server (ch ⟨ q′ ⟩ (∗-un (un₁ ⟨ p′ ⟩ un₂) , cut (↭proc swap P ⟨ (< p′) ⟩ server (ch ⟨ < r ⟩ (un₂ , Q)))))
  s-client :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (client (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
    client (ch ⟨ q′ ⟩ cut (↭proc swap P ⟨ < p′ ⟩ Q))
  s-weaken :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (weaken (ch ⟨ (> q) ⟩ P) ⟨ p ⟩ Q) ⊒ weaken (ch ⟨ q′ ⟩ cut (P ⟨ p′ ⟩ Q))
  s-contract :
    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut {A} (contract (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
    contract (ch ⟨ q′ ⟩ cut (↭proc (↭shift {A} {`? B ∷ `? B ∷ []}) P ⟨ < < p′ ⟩ Q))
\end{code}

@@

\begin{code}
  s-refl  : ∀{P} → P ⊒ P
  s-tran  : ∀{P Q R} → P ⊒ Q → Q ⊒ R → P ⊒ R
  s-cong  : ∀{Γ₁ Γ₂ A P Q P′ Q′} (p : Γ ≃ Γ₁ + Γ₂) →
            P ⊒ Q → P′ ⊒ Q′ → cut {A} (P ⟨ p ⟩ P′) ⊒ cut (Q ⟨ p ⟩ Q′)
\end{code}

### 2025-12-12 paper/Process.lagda.tex
revised congruence

\begin{code}
data Proc : Context → Set where
  link      : ∀{A} → ∀[ Ch A ∗ Ch (dual A) ⇒ Proc ]
  fail      : ∀[ Ch ⊤ ∗ U ⇒ Proc ]
  wait      : ∀[ Ch ⊥ ∗ Proc ⇒ Proc ]
  close     : ∀[ Ch 𝟙 ⇒ Proc ]
  case      : ∀{A B} →
              ∀[ Ch (A & B) ∗ ((A ∷_) ⊢ Proc ∩ (B ∷_) ⊢ Proc) ⇒ Proc ]
  select    : ∀{A B} →
              ∀[ Ch (A ⊕ B) ∗ ((A ∷_) ⊢ Proc ∪ (B ∷_) ⊢ Proc) ⇒ Proc ]
  join      : ∀{A B} → ∀[ Ch (A ⅋ B) ∗ ((A ∷_) ⊢ (B ∷_) ⊢ Proc) ⇒ Proc ]
  fork      : ∀{A B} →
              ∀[ Ch (A ⊗ B) ∗ ((A ∷_) ⊢ Proc) ∗ ((B ∷_) ⊢ Proc) ⇒ Proc ]
  all       : ∀{A} →
              ∀[ Ch (`∀ A) ∗ ⋂[ X ∶ Type ] ((subst [ X /_] A ∷_) ⊢ Proc) ⇒ Proc ]
  ex        : ∀{A B} → ∀[ Ch (`∃ A) ∗ ((subst [ B /_] A ∷_) ⊢ Proc) ⇒ Proc ]
  server    : ∀{A} → ∀[ Ch (`! A) ∗ (Un ∩ ((A ∷_) ⊢ Proc)) ⇒ Proc ]
  client    : ∀{A} → ∀[ Ch (`? A) ∗ ((A ∷_) ⊢ Proc) ⇒ Proc ]
  weaken    : ∀{A} → ∀[ Ch (`? A) ∗ Proc ⇒ Proc ]
  contract  : ∀{A} → ∀[ Ch (`? A) ∗ ((`? A ∷_) ⊢ (`? A ∷_) ⊢ Proc) ⇒ Proc ]
  cut       : ∀{A} → ∀[ ((A ∷_) ⊢ Proc) ∗ ((dual A ∷_) ⊢ Proc) ⇒ Proc ]
\end{code}

@@

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
$\Context$ satisfies the predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
a typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

We conclude this module proving that permutations preserve process typeability.
Since list permutations basically correspond to channel renaming, we can read
this property as the fact that typeability is preserved by (injective) name
substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
della separating conjunction introduce rumore, ma avere i costruttori
currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
questo rilievo nella prossima sezione quando è evidente il pattern matching che
serve per esprimere le regole.}

>>>>

\begin{code}
data Proc : Context → Set where
  link      : ∀{A} → ∀[ Ch A ∗ Ch (dual A) ⇒ Proc ]
  fail      : ∀[ Ch ⊤ ∗ U ⇒ Proc ]
  wait      : ∀[ Ch ⊥ ∗ Proc ⇒ Proc ]
  close     : ∀[ Ch 𝟙 ⇒ Proc ]
  case      : ∀{A B} →
              ∀[ Ch (A & B) ∗ ((A ∷_) ⊢ Proc ∩ (B ∷_) ⊢ Proc) ⇒ Proc ]
  select    : ∀{A B} →
              ∀[ Ch (A ⊕ B) ∗ ((A ∷_) ⊢ Proc ∪ (B ∷_) ⊢ Proc) ⇒ Proc ]
  join      : ∀{A B} → ∀[ Ch (A ⅋ B) ∗ ((A ∷_) ⊢ (B ∷_) ⊢ Proc) ⇒ Proc ]
  fork      : ∀{A B} →
              ∀[ Ch (A ⊗ B) ∗ ((A ∷_) ⊢ Proc) ∗ ((B ∷_) ⊢ Proc) ⇒ Proc ]
  all       : ∀{A} →
              ∀[ Ch (`∀ A) ∗ ⋂[ X ∶ Type ] ((subst [ X /] A ∷_) ⊢ Proc) ⇒ Proc ]
  ex        : ∀{A B} → ∀[ Ch (`∃ A) ∗ ((subst [ B /] A ∷_) ⊢ Proc) ⇒ Proc ]
  server    : ∀{A} → ∀[ Ch (`! A) ∗ (Un ∩ ((A ∷_) ⊢ Proc)) ⇒ Proc ]
  client    : ∀{A} → ∀[ Ch (`? A) ∗ ((A ∷_) ⊢ Proc) ⇒ Proc ]
  weaken    : ∀{A} → ∀[ Ch (`? A) ∗ Proc ⇒ Proc ]
  contract  : ∀{A} → ∀[ Ch (`? A) ∗ ((`? A ∷_) ⊢ (`? A ∷_) ⊢ Proc) ⇒ Proc ]
  cut       : ∀{A} → ∀[ ((A ∷_) ⊢ Proc) ∗ ((dual A ∷_) ⊢ Proc) ⇒ Proc ]
\end{code}

@@

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
$\Context$ satisfies the predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
a typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

We conclude this module proving that permutations preserve process typing. Since
list permutations basically correspond to channel renaming, we can read this
property as the fact that typing is preserved by (injective) name
substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
della separating conjunction introduce rumore, ma avere i costruttori
currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
questo rilievo nella prossima sezione quando è evidente il pattern matching che
serve per esprimere le regole.}
